MAESTROeX: A Massively Parallel Low Mach Number Astrophysical Solver

### 2019-08-02 MAESTROeX_I/appendix.tex
Some touchups
finished pass

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htb]
\begin{center}
\begin{tabular}{l c}
\multirow{4}{3.25in}[30mm]{ \includegraphics[width=3.0in]{./figs/wdconvect_256_splitU} } & \multicolumn{1}{c}{\includegraphics[width=2.35in]{./figs/magvel_full_XY}} \\
& \multicolumn{1}{c}{\begin{footnotesize} (a) Velocity magnitude, solved using full $\mathbf{U}$ \end{footnotesize}} \\[1.em]
& \multicolumn{1}{c}{\includegraphics[width=2.75in]{./figs/magvel_lineout_X}} \\ 
& \multicolumn{1}{c}{\begin{footnotesize} (b) Comparison along white line \end{footnotesize}} \\
\end{tabular}
\caption{\label{fig:wdconvect_splitU} (Left) Peak temperature, $T_{\text{peak}}$ in a white dwarf at resolution of $256^3$ 
         during initial transition. (Right) Velocity magnitude solved using full velocity $\mathbf{U}$ in the projection step 
         results in much larger values outside the density cutoff region, $\rho < 10^5 \text{ g cm}^{-3}$, that is not seen when 
         we split the velocity dynamics.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{description}
\item For planar geometry, $\etarho = \overline{\rho'(\Ub\cdot\eb_r)}$,
\begin{equation}
 \etarho^{\nph} =  {\rm {\bf Average}} \sum_k \left[ \left(\uadvtwo \cdot \eb_r \right) (\rho X_k)^{\nph,\pred} \right]
\end{equation}
\item For spherical geometry, first construct 
$\etarho^{{\rm cart},\nph} = [\rho'(\Ub\cdot\eb_r)]^{\nph}$ on Cartesian cell centers using:
\begin{equation}
\etarho^{{\rm cart},\nph} = \left[\left(\frac{\rho^{(1)}+\rho^{(2)}}{2}\right)-\left(\frac{\rho_0^n+\rho_0^{n+1}}{2}\right)\right] \cdot \left( \uadvtwo \cdot \eb_r \right).
\end{equation}
Then,
\begin{equation}
\etarho^{\nph,\star} = {\rm {\bf Average}}\left(\etarho^{{\rm cart},\nph}\right).
\end{equation}
\end{description}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[htb]
\begin{center}
\begin{tabular}{l c}
\multirow{4}{3.25in}[30mm]{ \includegraphics[width=3.0in]{./figs/wdconvect_256_splitU} } & \multicolumn{1}{c}{\includegraphics[width=2.35in]{./figs/magvel_full_XY}} \\
& \multicolumn{1}{c}{\begin{footnotesize} (a) Velocity magnitude, solved using full $\mathbf{U}$ \end{footnotesize}} \\[1.em]
& \multicolumn{1}{c}{\includegraphics[width=2.75in]{./figs/magvel_lineout_X}} \\ 
& \multicolumn{1}{c}{\begin{footnotesize} (b) Comparison along white line \end{footnotesize}} \\
\end{tabular}
\caption{\label{fig:wdconvect_splitU} When solving for the full velocity $\mathbf{U}$ in the nodal projection step instead of the split velocity 
  formulation, we illustrate
  (left) the peak temperature, $T_{\text{peak}}$ in a white dwarf at resolution of $256^3$ during the early evolution of the star and
  (right) the magnitude of velocity at early time. We observe much larger velocities outside the density cutoff region, $\rho < 10^5 \text{ g cm}^{-3}$, 
  that are not seen when we split the velocity dynamics.}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{description}
\item For planar geometry, $\etarho = \overline{\rho'(\Ub\cdot\eb_r)}$,
\begin{equation}
 \etarho^{\nph} =  {\rm {\bf Average}} \sum_k \left[ \left(\uadvtwo \cdot \eb_r \right) (\rho X_k)^{\nph,\pred} \right]
\end{equation}
\item For spherical geometry, first construct 
$\etarho^{{\rm cart},\nph} = [\rho'(\Ub\cdot\eb_r)]^{\nph}$ on Cartesian cell centers using:
\begin{equation}
\etarho^{{\rm cart},\nph} = \left[\left(\frac{\rho^{(1)}+\rho^{(2)}}{2}\right)-\left(\frac{\rho_0^n+\rho_0^{n+1}}{2}\right)\right] \cdot \left( \uadvtwo \cdot \eb_r \right).
\end{equation}
Then,
\begin{equation}
\etarho^{\nph} = {\rm {\bf Average}}\left(\etarho^{{\rm cart},\nph}\right).
\end{equation}
\end{description}

### 2019-08-02 MAESTROeX_I/results.tex
finished pass

To explore the accuracy of the new temporal algorithm, we now analyze in detail three-dimensional, full-star calculations of convection preceding ignition in a white dwarf. Again, we refer the reader to \cite{MAESTRO_AMR} and \cite{MAESTRO_convection} for setup details. We implement both uniformly- and irregularly-spaced base state with the new temporal algorithm, while only uniform base state spacing is used in the original algorithm. As in Section 3 of \cite{MAESTRO_AMR}, we choose the peak temperature and peak Mach number as the two diagnostics to compare the simulations. Figure \ref{fig:wdconvect_256_maxvar} shows the evolution of both peak temperature and peak Mach number until time of ignition on a single-level grid with resolution of $256^3$. The simulation using the new temporal scheme with uniformly-spaced base state gives the same qualitative results as the original scheme, and predicts a similar time of ignition ($t=7810$ s compared to $t=7850$ s for original algorithm). The simulation using the new temporal scheme with irregularly-spaced base state displays a slightly different peak temperature behavior during the initial transition period $t<150$ s, which results in the difference between the curves post transition. We strongly suspect that this is a result of using different initial model inputs for the star in which the resolution near the center of the star is much coarser with the irregular spacing than the uniform spacing. Fortunately, the simulation with irregular base state spacing still follows the same trend as with uniform spacing, and the star is shown to ignite at an earlier time $t=6840$ s.

Figure \ref{fig:wdconvect_Tmax} shows the peak temperature evolution over the first 1000 s on two grids of differing resolutions, $256^3$ and $512^3$. Limited allocations prevented us from running this simulation further.
As previously suspected, the simulation using irregularly-spaced base state agrees much closer to the results from using uniform spacing as the resolution is increased. This is most likely due to the increased resolution of the initial star model for the irregular-spaced base state that is now able to properly capture the same properties near the center of the star as its uniformly-spaced counterpart. This is especially important when computing the base state pressure from base state density since we enforce a constant base state pressure outside of a prescribed cutoff density radius that could potentially be quite different with the two types of base state spacing at lower resolutions.

@@

In terms of efficiency, all three simulations on $256^3$ single-level grid were run on Cori haswell with 64 processors and 8 threads per core and their run times were compared. As a result of simplifying the algorithm by eliminating the evolution equations for the base state density and pressure, the simulation using the new temporal algorithm took only 6.75 s per time step with uniformly-spaced base state, which is 13\% faster than the 7.77 s per time step when using the original scheme. However, we do observe a 25\% increase in run time of 9.72 s when using irregularly-spaced base state with the new algorithm. This can be explained by the irregularly-spaced base state array being much larger in size than its uniformly-spaced counterpart, and thus require additional computation time. One possible strategy to significantly reduce the run time is to consider truncating the base state beyond the cutoff density radius.

>>>>

To explore the accuracy of the new temporal algorithm, we now analyze in detail three-dimensional, full-star calculations of convection preceding ignition in a white dwarf. Again, we refer the reader to \cite{MAESTRO_AMR} and \cite{MAESTRO_convection} for setup details. We implement both uniformly- and irregularly-spaced base state with the new temporal algorithm, while only uniform base state spacing is used in the original algorithm. As in Section 3 of \cite{MAESTRO_AMR}, we choose the peak temperature and peak Mach number as the two diagnostics to compare the simulations. Figure \ref{fig:wdconvect_256_maxvar} shows the evolution of both peak temperature and peak Mach number until time of ignition on a single-level grid with resolution of $256^3$. The simulation using the new temporal scheme with uniformly-spaced base state gives the same qualitative results as the original scheme, and predicts a similar time of ignition ($t=7810$ s compared to $t=7850$ s for original algorithm). The simulation using the new temporal scheme with irregularly-spaced base state displays a slightly different peak temperature behavior during the initial transition period $t<150$ s, which results in the difference between the curves post transition. We strongly suspect that this is a result of using a different initial model file (the resolution near the center of the star is much coarser with the irregular spacing than the uniform spacing). Fortunately, the simulation with irregular base state spacing still follows the same trend as with uniform spacing, and the star is shown to ignite at an earlier time $t=6840$ s.

Figure \ref{fig:wdconvect_Tmax} shows the peak temperature evolution over the first 1000 s on two grids of differing resolutions, $256^3$ and $512^3$.
Limited allocations prevented us from running this simulation further.
As previously suspected, the simulation using irregularly-spaced base state agrees much closer to the results from using uniform spacing as the resolution increases.
This is most likely due to the increased resolution of the initial model, which more closely matches the uniformly-spaced counterpart.
This is especially important when computing the base state pressure from base state density, which is particularly sensitive to coarse resolution near the center of the star.
\MarginPar{AJN: Simplified this paragraph - OK?}

@@

In terms of efficiency, all three simulations on $256^3$ single-level grid were run on Cori haswell with 64 processors and 8 threads per core and their run times were compared. As a result of simplifying the algorithm by eliminating the evolution equations for the base state density and pressure, the simulation using the new temporal algorithm took only 6.75 s per time step with uniformly-spaced base state, which is 13\% faster than the 7.77 s per time step when using the original scheme. However, we do observe a 25\% increase in run time of 9.72 s when using irregularly-spaced base state with the new algorithm. This can be explained by the irregularly-spaced base state array being much larger in size than its uniformly-spaced counterpart, and thus require additional communication and computation time. One possible strategy to significantly reduce the run time is to consider truncating the base state beyond the cutoff density radius.
